The imprint of global magma oceans on exoplanet demographics

### 2023-04-09 src/tex/ms.tex
Updated Section 3

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
\todo[inline]{TIM: please review}
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019}.
We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience.
...

@@

\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
This case serves as our null hypothesis.

For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius change due to molten interiors~\citep{Bower2019}.
It further assumes a \textit{dry melt} without dissolved volatiles.

\textit{Wet melt} planets follow the rules above, but we take into account a radius decrease from water partitioning in the melt:
In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
This removes it from the steam atmosphere, which in turn decreases the radius inflation.
The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes are predicted to reach measurable values~\citep{Dorn2021}.
The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.

\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
    \end{centering}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiusevolution.pdf}
        \caption{
            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states. Green: planets with a solid mantle and no steam atmosphere. Dashed: planets with steam atmospheres. Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
                Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes. Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios. Planets can move between planet classes through processes such as atmospheric loss and desiccation, which ultimately ends a runaway greenhouse phase on a timescale dependent on a planet's water content.}
        \label{fig:radiusevolution}
    \end{centering}
\end{figure}
We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed. % Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.

To account for planets unable to sustain a steam atmosphere as well as evolutionary effects such as desiccation (see bottem panel of Fig.~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
%The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.

@@

\subsubsection{Simplifications in the atmosphere model}
\todo[inline]{TIM: your expertise will be valuable here.\\
Mention we only consider instellation-induced rgh.\\
- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
}
As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the runaway or non-runaway category.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the non-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.

>>>>

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{TIM: please review} DONE
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}. With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state \citep{2021JGRE..12606711L,2021ApJ...919..130B}. We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean \citep{Dorn2021,2023Icar..39015265S}.

@@

\textit{Sub-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
This case serves as our null hypothesis.

\textit{Super-runaway} planets: For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,2013ApJ...765..131K,Leconte2013,2015ApJ...806..216H,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our sub-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius increase due to the density deficit of molten interiors~\citep{Bower2019}.

Nominally, the above models assume a \textit{dry melt} without dissolved volatiles. Here, however, we consider a \textit{wet melt} magma ocean, take into account a radius decrease from water partitioning in the melt, following \citet{Dorn2021}: In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
This removes it from the steam atmosphere, which in turn decreases the radius inflation effect.
The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes are predicted to reach measurable values~\citep{Dorn2021}.
The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.

\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
    \end{centering}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiusevolution.pdf}
        \caption{
            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states. Green: planets with a solid mantle and no steam atmosphere. Dashed: planets with steam atmospheres. Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
                Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes. Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios. Planets can move between planet classes through processes such as atmospheric loss and desiccation, which ultimately ends a runaway greenhouse phase on a timescale dependent on a planet's water content.}
        \label{fig:radiusevolution}
    \end{centering}
\end{figure}
We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed. % Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
In the following, we only distinguish between the sub-runaway greenhouse and wet melt scenarios.

To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation (see bottem panel of Fig.~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the sub-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation. The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.

@@

\subsubsection{Simplifications in the atmosphere model}
\todo[inline]{TIM: your expertise will be valuable here.\\
Mention we only consider instellation-induced rgh.\\
- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
}
As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the runaway or sub-runaway category.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
